Simulation

### 2023-12-04 simulation.qmd
Init Quarto

---
title: "Simulation"
execute: 
  cache: true
format: 
  html:
    toc: true
    code-fold: true
---

```{r}
#| label: options

library(posterior)
library(mvtnorm)
library(cmdstanr)
library(ggdist)
library(brms)
library(tidyverse)

@@

In this document we specify, simulate, and validate a univariate latent means brms model. Code based on Joran's blog post at <https://experienced-sampler.netlify.app/post/stan-hierarchical-ar/>.

# Data

We first set up the data simulation. I used [this code](https://experienced-sampler.netlify.app/post/stan-hierarchical-ar/#testing) but put it inside a function.

@@

## Stan model

Let us then compile and sample from the cmdstanr model

```{r}
# model <- cmdstan_model("model-ar1.stan", compile = FALSE)
# model$format(
#   canonicalize = list("deprecations"), 
#   overwrite_file = TRUE, 
#   backup = FALSE
# )
# model$compile()
model <- cmdstan_model(exe_file = "model-ar1")
path <- "models/cmdstanr-example.rds"
if (!file.exists(path)) {
post <- model$sample(
  data = dat, 
  adapt_delta = 0.99
) |> 
  as_draws_df()
saveRDS(post, path)
} else {post <- readRDS(path)}
```

Visualize estimated and defined parameters

```{r}
post <- post |> 
  select(starts_with("beta_"), ends_with("_raw")) |> 
  rename_with(~str_remove(., "_raw"))
post |> 
  pivot_longer(everything()) |> 
  ggplot(aes(value, name)) +
  scale_x_continuous(
    "Parameter value"
  ) +
  stat_histinterval(
    normalize = "xy",
    scale = 0.25, 
    size = 1
  ) +
  geom_point(
    data = as_tibble(dat) |> 
      select(sigma:beta_scale) |> 
      distinct() |> 
      pivot_longer(everything()),
    position = position_nudge(y = -0.1),
    color = "red"
  ) +
  theme(
    axis.title.y = element_blank()
  )
```

## brms

```{r}
dat <- as_tibble(dat)
dat <- dat |> 
  mutate(yt1 = lag(y), .by = individual)
dat
```

>>>>

---
title: Simulation
subtitle: Specification and parameter recovery simulation of brms latent means model
execute: 
  error: true
  message: false
  warning: false
---

Here we simulate from, specify, and validate latent means models with brms.

```{r echo = FALSE}
#| label: options
#| output: false

library(posterior)
library(mvtnorm)
library(scales)
library(gt)
library(cmdstanr)
library(ggdist)
library(future)
library(brms)
library(tidyverse)

@@

## Univariate model

The data simulation and Stan code here is based on [Joran's blog post](https://experienced-sampler.netlify.app/post/stan-hierarchical-ar/).

The model is for each individual i and timepoint t,

y_it ~ normal(eta_it, sigma)
eta_it = (alpha_hat + alpha_i) + (beta_hat + beta_i) * (y_i[t-1])
[alpha_i, beta_i] ~ mvn([0, 0], Sigma)

That is we model y on an intercept and an autoregressive coefficient with varying intercepts and slopes over people.

### Data

We first set up the data simulation.

@@

### Stan

Let us then compile and sample from the Stan model (`model-ar1.stan`) with cmdstanr

```{r}
#| output: false

path <- "cache/cmdstanr-example.rds"
if (!file.exists(path)) {
  model <- cmdstan_model("model-ar1.stan", compile = FALSE)
  model$format(
    canonicalize = list("deprecations"),
    overwrite_file = TRUE,
    backup = FALSE
  )
  model$compile()
  post_cmdstanr <- model$sample(
    data = dat, 
    adapt_delta = 0.99
  ) |> 
    as_draws_df()
  saveRDS(post_cmdstanr, path)
} else {
  post_cmdstanr <- readRDS(path)
}
```

Posterior summary:

```{r}
post_cmdstanr <- post_cmdstanr |> 
  select(
    alpha_hat_raw, alpha_scale_raw, 
    beta_hat, beta_scale, 
    sigma_raw
  ) |> 
  rename_with(~str_remove(., "_raw"))

summarize_draws(post_cmdstanr) |> 
  gt() |> 
  fmt_number(decimals = 2)
```

### brms

brms takes a data frame, so we convert the list here

```{r}
dat <- as_tibble(dat)
dat <- dat |> 
  mutate(yt1 = lag(y), .by = individual)
head(dat) |> 
  gt() |> 
  fmt_number(decimals = 2)
```

#### Model specification

We can specify this model using brms' nonlinear formulas. Note that we created the `yt1` variable (i.e. y[t-1]) above.

```{r}
#| output: false

bform <- bf(
  y ~ 0 + alpha + beta * (yt1 - alpha),
  alpha ~ 1 + (1 | individual),
  beta ~ 1 + (1 | individual),
  nl = TRUE
)
fit_brms <- brm(
  bform,
  data = dat,
  file = "cache/brm-example.rds"
)
```

We see that the estimated parameters are very close to coefficients from the Stan model

```{r}
post_brms <- as_draws_df(fit_brms) |> 
  select(
    alpha_hat = b_alpha_Intercept,
    alpha_scale = sd_individual__alpha_Intercept,
    beta_hat = b_beta_Intercept,
    beta_scale = sd_individual__beta_Intercept,
    sigma
  )

bind_rows(
  "Stan" = summarize_draws(post_cmdstanr),
  "brms" = summarize_draws(post_brms),
  .id = "Model"
) |> 
  arrange(variable, Model) |> 
  gt() |> 
  fmt_number(decimals = 2)
```

### Simulation

Above, we estimated the Stan and brms models once. Here, we evaluate whether they recover the specified parameters with a small simulation.

```{r}
path <- "cache/sim-univariate.rds"
if (!file.exists(path)) {
  # Ensure that models are compiled
  model_stan <- cmdstan_model("model-ar1.stan", compile = TRUE)
  fit_brms <- brm(
    bform,
    data = dat,
    chains = 0,
    threads = 1
  )

  # Define simulation outputs
  K <- 30
  out <- vector("list", length = K)

  # Parallelization scheme:
  # Each of two workers will launch a session that runs 4 chains in parallel
  # resulting in total 8 cores used
  plan(multisession, workers = 2)

  for (k in 1:K) {
    this_sample <- simulate()

    this_stan <- future({
      model_stan$sample(
        data = this_sample, 
        adapt_delta = 0.99, 
        parallel_chains = 4,
        iter_sampling = 2000,
        refresh = 0, 
        show_messages = FALSE
      )
    }, seed = TRUE)

    this_brms <- future({
      update(
        fit_brms, 
        newdata = as_tibble(this_sample) |> 
          mutate(yt1 = lag(y), .by = individual),
        control = list(adapt_delta = 0.99),
        cores = 4,
        iter = 2000,
        recompile = FALSE,
        refresh = 0,
        silent = 2
      )
    }, seed = TRUE)

    this_stan <- value(this_stan) 
    this_stan <- this_stan |> 
      as_draws_df() |> 
      select(
        alpha_hat_raw, alpha_scale_raw, 
        beta_hat, beta_scale, 
        sigma_raw
      ) |> 
      rename_with(~str_remove(., "_raw")) |> 
      summarize_draws()

    this_brms <- value(this_brms) 
    this_brms <- this_brms |> 
      as_draws_df() |> 
      select(
        alpha_hat = b_alpha_Intercept,
        alpha_scale = sd_individual__alpha_Intercept,
        beta_hat = b_beta_Intercept,
        beta_scale = sd_individual__beta_Intercept,
        sigma
      ) |> 
      summarize_draws()

    out[[k]] <- bind_rows(
      "Stan" = this_stan,
      "brms" = this_brms,
      .id = "model"
    )
  }

  out <- bind_rows(out, .id = "k")
  saveRDS(out, path)
} else {out <- readRDS(path)}
```

#### Results

```{r}
#| label: tbl-simulation

real_values <- tribble(
  ~variable, ~true,
  "sigma", 1, 
  "alpha_hat", 4, 
  "alpha_scale", 1, 
  "beta_hat", 0.4, 
  "beta_scale", 0.1
)
out |> 
  summarise(
    mean = str_glue("{number(mean(mean), .01)} ({number(sd(mean), .01)})"),
    .by = c(model, variable)
  ) |> 
  pivot_wider(
    names_from = model, values_from = mean
  ) |> 
  left_join(
    real_values
  ) |> 
  gt() |> 
  fmt_number(decimals = 1) |> 
  fmt_missing(missing_text = " ")
```

## Conclusion

This small simulation shows that brms perfectly recovers parameters of a simple univariate autoregressive latent variable model. The mean estimates over 30 simulation iterations were identical for brms and Stan to the second significant digit, and within MCSE of the true values.
